---
title: "Quick start guide"
author: "Martin Westgate"
date: '2024-05-10'
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`galaxias` is an R package to help users build repositories that are optimised
for storing, documenting and sharing biodiversity data. These repositories can 
be seamlessly converted into a **Darwin Core Archive**, the data standard used 
by the [Global Biodiversity Information Facility (GBIF)](https://gbif.org) and
it's partner nodes.


## Start a new project

To start a new data project using galaxias, call: 

```{r, eval=FALSE}
library(galaxias)
galaxias_project("myprojectname")
```

If you are using RStudio, this should launch a new RStudio instance for
that project, which will have the following structure:

```
├── README.md                        : Description of the repository
├── metadata.md                      : Boilerplate metadata statement for this project
├── projectname.Rproj                : RStudio project file
├── data-raw                         : Folder to store source data
└── data                             : Folder to store processed data
```

It is a good idea to write some basic information in your `README.Rmd` file
first, as this provides guidance to your users as to what your package contains,
as well as what they are allowed to use it for.

At this point, we suggest storing your repository on GitHub or similar. See ...
for advice on setting this up.

## Adding data to your project

We recommend that you first add your data to the `data-raw` folder, then use a 
script within that folder to manipulate it to a suitable interchange format 
(such as Darwin Core). Let's assume, for example, that your data is looks like 
this:

```{r}
library(tibble)

df <- tibble(
  latitude = c(-35.310, -35.273),
  longitude = c(149.125, 149.133),
  date = c("14-01-2023", "15-01-2023"),
  time = c("10:23", "11:25"),
  species = c("Callocephalon fimbriatum", "Eolophus roseicapilla"))
```

(Note that normally you'd import your data from an external file (e.g. using
`readr::read_csv()`), but we've constructed one here for example purposes.)

We recommend using the `corella` package for converting tibbles to Darwin 
Core. A minimally complete set of individual observations formatted using
this package might look like this:

```{r}
library(corella)
library(lubridate)

occurrences <- df |>
  # basic requirements of Darwin Core
  use_occurrences(occurrenceID = seq_len(nrow(df)),
                  basisOfRecord = "humanObservation") |> 
  # place and time
  use_coordinates(decimalLatitude = latitude, 
                  decimalLongitude = longitude) |>
  use_locality(country = "Australia", 
               locality = "City of Town") |>
  use_datetime(eventDate = lubridate::dmy(date),
               eventTime = lubridate::hm(time)) |>
  # taxonomy
  use_scientific_name(scientificName = species, 
                      scientificNameRank = "species") |>
  use_taxonomy(kingdom = "Animalia", 
               phylum = "Aves") 
```

Note that this deliberately includes some redundancy. The coordinate data are
useful by themselves, for example, but in case of ambiguity it is useful to
specify a text string given some information on location. Similarly, the 
`scientificName` field should be sufficient to identify the taxon in question, 
but adding higher taxonomic information makes the identification less ambiguous.

Although it is possible to provide data as a set of unrelated observations,
this structure doesn't reflect the way in which ecological surveys are typically
conducted. If, for example, you have conducted a survey in which many species
can be detected at once, you can try:

(add code block here showing `use_events()` and `use_abundance()`)

(add example of refining only to DwC terms)

We have saved our tibble as an object with a Darwin Core-specific name, so we 
can save it out using `write_csv()`

```{r, eval=FALSE}
library(readr)
write_csv(occurrences, ".data/occurrences.csv")
```

If your data are _already_ in Darwin Core format, you can simply place them 
in the data folder and remove your `data-raw` folder.

## Adding package metadata

Darwin Core Archives use `xml` files to store two metadata files; one that 
describes your data and ownership, and a second that stores its' structure.
We recommend using the `paperbark` package to manipulate these files.

`build_schema()` example

`build_metadata()` example

Finally, best to `build_archive()`.

## Check your respository

Darwin Core may be an unfamiliar format, so it can be useful to 'check' your
data for common issues. We suggest first using `check_archive()`:


Alternatively, you can use the GBIF 'validate' API to check your data (not functional!)

```{r, eval=FALSE}
validate_archive()
```

## Sharing data

