---
title: "Quick start guide"
author: "Martin Westgate"
date: '2024-12-12'
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`galaxias` is an R package to help users build repositories that are optimised
for storing, documenting and sharing biodiversity data. These repositories can 
be seamlessly converted into a **Darwin Core Archive**, the data standard used 
by the [Global Biodiversity Information Facility (GBIF)](https://gbif.org) and
it's partner nodes.

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# load packages now to avoid messages later
library(galaxias)
library(lubridate)
```

## Start a new project

To start a new data project using galaxias, call: 

```{r, eval=TRUE, echo=FALSE}
devtools::load_all()
```

```{r, eval=FALSE}
library(galaxias)
galaxias_project("my-project-name")
```

If you are using RStudio, this should launch a new RStudio instance for
that project, which will have the following structure:

```
├── README.md                        : Description of the repository
├── metadata.md                      : Boilerplate metadata statement for this project
├── my-project-name.Rproj            : RStudio project file
├── data-raw                         : Folder to store source data
└── data                             : Folder to store processed data
```

It is a good idea to write some basic information in your `README.Rmd` file
first, as this provides guidance to your users as to what your package contains,
as well as what they are allowed to use it for.

## Adding data to your project

We recommend that you first add your data to the `data-raw` folder, then use a 
script within that folder to manipulate it to Darwin Core format. Let's assume, 
for example, that your data is looks like this:

```{r}
library(tibble)

my_tbl <- tibble(
  latitude = c(-35.310, -35.273),
  longitude = c(149.125, 149.133),
  date = c("14-01-2023", "15-01-2023"),
  time = c("10:23", "11:25"),
  species = c("Callocephalon fimbriatum", "Eolophus roseicapilla"))
```

(Note that normally you'd import your data from an external file (e.g. using
`readr::read_csv()`), but we've constructed one here for example purposes.)

You can convert tibbles to use Darwin Core standard using `set_` functions 
imported from the [corella package](https://corella.ala.org.au/) 
(automatically loaded with galaxias). A minimally complete set of formatted 
observations might look like this:

```{r, echo=FALSE, message=FALSE}
library(lubridate)

occurrences <- my_tbl |>
  # basic requirements of Darwin Core
  set_occurrences(occurrenceID = sequential_id(),
                  basisOfRecord = "humanObservation") |> 
  # place and time
  set_coordinates(decimalLatitude = latitude, 
                  decimalLongitude = longitude) |>
  set_locality(country = "Australia", 
               locality = "Canberra") |>
  set_datetime(eventDate = lubridate::dmy(date),
               eventTime = lubridate::hm(time)) |>
  # taxonomy
  set_scientific_name(scientificName = species, 
                      taxonRank = "species") |>
  set_taxonomy(kingdom = "Animalia",
               phylum = "Aves") 
```

```{r, echo=TRUE, eval=FALSE}
library(lubridate)

occurrences <- df |>
  # basic requirements of Darwin Core
  set_occurrences(occurrenceID = sequential_id(),
                  basisOfRecord = "humanObservation") |> 
  # place and time
  set_coordinates(decimalLatitude = latitude, 
                  decimalLongitude = longitude) |>
  set_locality(country = "Australia", 
               locality = "Canberra") |>
  set_datetime(eventDate = lubridate::dmy(date),
               eventTime = lubridate::hm(time)) |>
  # taxonomy
  set_scientific_name(scientificName = species, 
                      taxonRank = "species") |>
  set_taxonomy(kingdom = "Animalia", 
               phylum = "Aves") 
```

```{r, echo=TRUE}
print(occurrences, n = 5)
```

Note that this deliberately includes some redundancy. The coordinate data are
useful by themselves, for example, but in case of ambiguity it is useful to
specify a text string given some information on location. Similarly, the 
`scientificName` field should be sufficient to identify the taxon in question, 
but adding higher taxonomic information makes the identification less ambiguous.

We have saved our tibble as an object with a Darwin Core-specific name, so we 
can subset to only Darwin Core terms, and save it out using `write_csv()`

```{r, eval=FALSE}
library(readr)
occurrences |>
  select(any_of(occurrence_terms())) |>
  use_data_occurrences(".data/occurrences.csv")
```

If your data are _already_ in Darwin Core format, you can simply place them 
in the `data` folder. You can then use `build_schema()` to create a 'schema'
file, which is an `xml` document that tells users what data is present in your 
archive.

## Adding package metadata

A critical part of a Darwin Core archive is a metadata statement; this tells 
users who owns the data, what the data were collected for, and what uses they
can be put to (i.e. a data licence). To get an example statement, call
`use_metadata()`

```{r, eval=FALSE}
use_metadata()
```

This creates a blank statement called 'metadata.md', which looks like this:

```{r, echo=FALSE, eval=FALSE}
# this code doesn't work any more
# best practice here might be to call `use_metadata()` then `readLines()` and `cat()`
library(delma)
metadata_string <- as_eml_chr(metadata_example)[3:15]
  metadata_string |>
  paste0("\n") |>
  cat()
```

Once you have editted the statement to reflect the information you want to
convey, you can convert it to `EML` using:

```{r, eval=FALSE}
use_metadata()
```

A second piece of metadata required by the Darwin Core standard is the 'schema'
document. This is a machine-readable `xml` document that describes the content
of the archive's data files. You can generate one using:

```{r, eval=FALSE}
build_schema()
```

## Build an archive
At the end of the above process, you should have a folder named `data` that 
contains at least three files:

- One or more `.csv` files containing data
- a `meta.xml` file containing your schema
- an `eml.xml` file containing your metadata

If that is true, then you can run `build_archive()` to zip your data into a 
Darwin Core Archive:

```{r eval=FALSE}
build_archive()
```
